Inferring community characteristics in labelled networks

### 2021-05-16 nips/abstract.tex
Work on intro

	Nevertheless, these feature-based communities are often not a natural partition of the graph and thus the model is not a good fit. With this in mind, we present a novel generative model for better describing vertex-labelled undirected graphs. This is a two-stage model. First, we start with vertex features and probabilistically generate a block membership. Secondly, given these latent block memberships, we draw a graph according to the micro-canonical stochastic block model. We call this model the latent block generative model (LBGM).

	Armed with the newly defined LBGM, we present a method to efficiently sample the parameters of the feature to block generator. This allows us to automatically determine which features best explain the block-based graphical structure. Importantly, this analysis can be performed using the whole feature-set rather than considering each independently.

@@

	Future work may benefit from extending the LBGM to multiple hierarchical levels. Currently, the single-level model only provides explanations for the widest separation between communities. Nevertheless, in real-world data each community may be composed of many sub-communities and so on. A hierarchical approach would be able to find explanations for network structure at each level of the hierarchy. 
\end{abstract}

>>>>

	Nevertheless, these feature-based communities are often not a natural partition of the graph and thus the model is not a good fit. With this in mind, we present a novel generative model for better describing vertex-labelled undirected graphs. This is a two-stage model. First, we start with vertex features and probabilistically generate a block membership. Secondly, given these latent block memberships, we draw a graph according to the micro-canonical stochastic block model. We call this model the feature-first block model (FFBM).

	Armed with the newly defined FFBM, we present a method to efficiently sample the parameters of the feature to block generator. This allows us to automatically determine which features best explain the block-based graphical structure. Importantly, this analysis can be performed using the whole feature-set rather than considering each independently.

@@

	Future work may benefit from extending the FFBM to multiple hierarchical levels. Currently, the single-level model only provides explanations for the widest separation between communities. Nevertheless, in real-world data each community may be composed of many sub-communities and so on. A hierarchical approach would be able to find explanations for network structure at each level of the hierarchy. 
\end{abstract}

### 2021-05-16 nips/inference.tex
Work on intro

Now that we have defined the LBGM, we wish to leverage it to perform inference. Suppose we are presented with a vertex-labelled graph $(A, X)$; the goal is to draw samples for $\theta$ according to the posterior given the observed graph (equation \ref{eqn:theta-target}). 
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(t)} \sim p(\theta | A, X)
\end{equation}
%
These samples allow us to approximate the posterior distribution for $\theta$ as well as compute a predictive distribution $p(b^* | x^*, A, X) = \int p(b^* | x^*, \theta) p(\theta | X, A) d\theta \approx \frac{1}{T} \sum_{t=1}^{T} p(b^* | x^*, \theta^{(t)})$. However, generating these samples is not easily done in practice.

@@

\begin{algorithm} % enter the algorithm environment
	\caption{LBGM parameter pseudo-marginal inference} % give the algorithm a caption
	\label{alg:theta-samples} % and a label for \ref{} commands later in the document
	\begin{algorithmic} % enter the algorithmic environment

@@

For the purposes of our model (the LBGM), the number of blocks $B$ is a constant which must be specified by the data scientist. We could however, allow our choice of $B$ to be influenced by the observed data. This places us in the domain of empirical Bayes, which must be negotiated carefully. Prior beliefs must be determined a priori else they are not prior. However, as the number of blocks only specifies the coarseness of the analysis, it is fine to allow it to vary. Indeed, \citet{peixoto-determine-B} shows that for a fixed average degree the maximum number of detectable blocks scales as $O(\sqrt{N})$ where $N$ is the number of vertices.

>>>>

Now that we have defined the FFBM, we wish to leverage it to perform inference. Suppose we are presented with a vertex-labelled graph $(A, X)$; the goal is to draw samples for $\theta$ according to the posterior given the observed graph (equation \ref{eqn:theta-target}). 
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(t)} \sim p(\theta | A, X)
\end{equation}
%
These samples allow us to approximate the posterior distribution for $\theta$ as well as compute a predictive distribution $p(b^* | x^*, A, X) = \int p(b^* | x^*, \theta) p(\theta | X, A) d\theta \approx \frac{1}{T} \sum_{t=1}^{T} p(b^* | x^*, \theta^{(t)})$. However, generating these samples is not easily done in practice.

@@

\begin{algorithm} % enter the algorithm environment
	\caption{FFBM parameter pseudo-marginal inference} % give the algorithm a caption
	\label{alg:theta-samples} % and a label for \ref{} commands later in the document
	\begin{algorithmic} % enter the algorithmic environment

@@

For the purposes of our model (the FFBM), the number of blocks $B$ is a constant which must be specified by the data scientist. We could however, allow our choice of $B$ to be influenced by the observed data. This places us in the domain of empirical Bayes, which must be negotiated carefully. Prior beliefs must be determined a priori else they are not prior. However, as the number of blocks only specifies the coarseness of the analysis, it is fine to allow it to vary. Indeed, \citet{peixoto-determine-B} shows that for a fixed average degree the maximum number of detectable blocks scales as $O(\sqrt{N})$ where $N$ is the number of vertices.

### 2021-05-16 nips/introduction.tex
Work on intro

There is a wealth of networks in the world. Some examples of this graphical data are social networks, website hyperlinks and academic collaboration with more being produced each second. It is clear we need tools to analyse this increasingly omni-present form of data.

A somewhat surprising property of real-world graphs is that they exhibit strong community structure. In other words, each node will often belong to a cluster of densely connected nodes. This property is often exploited by graph compression algorithms and there is high interest in recovering the communities from the observed graph.

A common subset of graphical data is the labelled network. This is a graph where we have information about the properties of each node. We shall refer to these node properties as features. One of the most common questions we can ask of this dataset is what features have the largest impact on the structure of the graph. For example, when analysing an academic collaboration graph, one may wish to ask what impact gender has on the structure of the network.  Nevertheless, this analysis is often vulnerable to confounding variables. While gender very well may impact the structure of the graph, there is often a better explanatory variable for the structure.

There is space to bridge the gap between these two approaches. Rather than determining whether a feature impacts the graphical structure directly, we can use the community concept as a stepping stone in our analysis. Therefore, we extract which features have the largest impact on overall graphical structure. This can be thought of as a form of dimensionality-reduction where we only keep the node features that are useful in predicting community memberships.

>>>>

A somewhat surprising property of many real-world networks is that they exhibit strong community structure. In other words, each node will often belong to a cluster of densely connected nodes. There is high interest in recovering the latent communities from the observed graphs. The inferred communities can be exploited for compression algorithms \cite{cluster-compression} or used for link prediction in incomplete networks \cite{link-prediction} to name but a few applications.

Nevertheless, we restrict our analysis to labelled networks. These are graphs where we have information about the properties of each node. We shall refer to these node properties as features. One of the most common questions we can ask of labelled networks is whether a given vertex feature has an impact on structure of the graph. To answer this question from a Bayesian perspective we must use a random graph model; the most common form is called the stochastic block model (SBM) \cite{vanilla-sbm}. This is a latent variable model where each vertex belongs to a single block and the probability two nodes are connected depends only on the. There have been many variants to this model; the most popular are the mixed-membership stochastic block model (MMSBM) \cite{mixed-membership-sbm} and overlapping stochastic block model (OSBM) \cite{overlapping-sbm}. Effectively, these just extend the model to allow each vertex to belong to multiple blocks simultaneously.

However, a major drawback of these graphical models as applied to labelled networks is that they do not automatically include vertex features in the random graph generation process. Approaches based on graph neural networks \cite{sbm-gnn} that utilise vertex features have been developed but these lack the easy interpretability of the simpler models. To analyse a labelled network using one of the simple SBM variants, a typical inference procedure would be to partition the graph into blocks grouped by distinct values of the feature of interest. The associated model can then be used to test for evidence of heterogeneous connectivity between the feature-grouped blocks. Nevertheless, this approach is limited in that it can only consider one feature at a time. Furthermore, it is hard to rank order the magnitude of the impact of each feature if they are considered independently. We would instead prefer to partition the graph into its most natural blocks and then determine which feature best predict the resulting partition.

With these desiderata in mind, we present a novel model for modelling labelled networks, which we call the feature-first block model (FFBM). This allows us to use the latent block membership as a stepping stone in our analysis. Therefore, we extract which features have the largest impact on overall graphical structure. This can be thought of as a form of dimensionality-reduction where we only keep the node features that are useful in predicting community memberships.
